---
title: 'HW 08: Control Flow'
author: "Chirag Kulkarni"
date: "3/2/2021"
output:
  html_document:
    toc: true
    toc_depth: 4
    number_sections: false
    toc_float:
      collapsed: false
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(magrittr)
```

## Question 1
Use the ifelse function to return “GROWING” if the elements of a vector x, with length 3, are strictly increasing, and “NOT growing” otherwise.
```{r}
#sets the three elements of the vector "x"
x <- c(11,22,33)
#uses the ifelse function to check if the elements are strictly increasing
growth <- ifelse(x[2]>x[1] & x[3]>x[2], "GROWING", "NOT growing")
#prints "growth" to check
growth
```


## Question 2
Use an else-if sequence to return “GROWING” if the elements of a vector x with length 3 are strictly increasing, “SHRINKING” if the elements of a vector x with length 3 are strictly decreasing, and “PEAKING” otherwise. Store the result in a variable named growth_pattern.

Then, use the switch() function to return the vector c("small", "Medium", "LARGE") if growth_pattern is "GROWING", return the factor factor(c("HUGE", "Teeny", "tiny")) if growth_pattern is "SHRINKING", and return the list list("valley", "PEAK", "valley") otherwise. Assign the result of switch size_pattern.
```{r}
#sets an initial vector with length 3
x <- c(11,22,33)
growth_pattern<-
#if the elements of the vector are growing, print "GROWING"
if (x[2]>x[1] & x[3]>x[2]){
  print("GROWING")
#if the elements of the vector are shrinking, print "SHRINKING"
}else if (x[1]>x[2] & x[2]>x[3]){
  print("SHRINKING")
#if the elements of the vector are doing anything else, print "PEAKING"
}else{
  print("PEAKING")
}
#prints growth pattern to check
growth_pattern
size_pattern<-
switch(growth_pattern,
       #if the elements are growing, print: "small", "Medium", "LARGE"
       "GROWING" = c("small", "Medium", "LARGE"),
       #if the elements are shrinking, print the factor: "HUGE", "Teeny", "tiny"
       "SHRINKING" = factor(c("HUGE", "Teeny", "tiny")),
       #if the elements are doing anything else print the list: "valley", "PEAK", "valley"
       "PEAKING"=list("valley", "PEAK", "valley")
       )
#prints size_pattern to check
size_pattern
```


## Question 3
Create a control sequence that, given a numeric vector x of length 3, will print the elements by order from high to low.
```{r}
#sets an initial vector with length 3
x <- c(17, 3, 8)

#if the initial first element is greater than the initial second element, than 
#the initial first element comes first and the initial second element comes second
if (x[1] > x[2]){
    fir <- x[1]
    sec <- x[2]
#else the initial second element comes first and the initial first element comes second
} else {
    fir <- x[2]
    sec <- x[1]
}
#if the initial third element is greater than the designated first element and
#the designated second element, the designated second element becomes the third designated 
#element, the designated first element becomes the designated second element, and the 
#third element becomes the designated first element
if ( x[3] > fir & x[3] > sec ) {
    thi <- sec
    sec <- fir
    fir <- x[3]
#else if the third element is less than the first and second designated elements,
#then the third element stays as the designated third element.
} else if ( x[3] < fir & x[3] < sec ) {
    thi <- x[3]
#else the designated second element becomes the designated third element and the 
#third element becomes the designated second element
} else {
    thi <- sec
    sec <- x[3]
}
#concatenates and prints the respective first, second, and third elements
cat (fir, sec, thi, "\n")
```


## Question 4
Write a for loop that iterates over the column names of the inbuilt iris dataset and print each, together with the number of characters in the column name in parentheses.

Example output: Sepal.Length (12).

Use the following functions, and any others you wish: paste0() and nchar().
```{r}
#this is a for loop that iterates over the column names of the inbuilt iris dataset 
#and print each, together with the number of characters in the column name in parentheses.
for (n in names(iris)) {
  print(paste0(n, " (", nchar(n), ")"))
}
```


## Question 5
Use a nested for loop (a for loop inside a for loop) that produces the following matrix, preallocate the matrix with NA values to a variable named row_minus_col.
```{r}
#creates a 5 by 5 matrix to set our conditions in
row_minus_col <- matrix(NA_integer_, nrow = 5, ncol = 5)
#creates a nested for loop that creates the given matrix
for (i in 1:5) {
  for (j in 1:5) {
    row_minus_col[i, j] <- i-j
  }
}
#prints out the final result
row_minus_col
```


## Question 6
Use a nested for loop (a for loop inside a for loop) that produces the following matrix, preallocate the matrix with NA values to a variable named row_col_distance.
```{r}
#creates a 5 by 5 matrix to set our conditions in 
row_col_distance <- matrix(NA_integer_, nrow = 5, ncol = 5)
#creates a nested for loop that creates the given matrix
for (i in 1:5) {
  for (j in 1:5) {
    row_col_distance[i, j] <- abs(i - j)
  }
}
#prints out the final result
row_col_distance
```


## Question 7
Use a while loop to investigate the number of terms required before the product
1⋅2⋅3⋅4⋅…
reaches above 1 million.
```{r}
#sets the target to reach as one million
target <- 1000000
#sets the initial product value to 1
product <- 1
#sets the count to 0
n <- 0
#fidns the number of terms requred before the product reaches one million
while (product <= target) {
  n <- n + 1
  product <- product*n
}
#prints the final count
n
```


## Question 8
Implement a multiplication game, a repeat loop (4 pts) that gives the user two random numbers from 2 to 12 (2 pts). Be sure the loop:

• Prints a message that asks the user to produce the product of the two numbers (include the actual numbers in the message). (2 pts)
• Takes their response as input using readline(). (2 pts)
• If their answer is correct:
    • Adds one to the correct answers count. (2 pts)
    • Let’s them know the answer is right. (2 pts)
    • Breaks if the number of correct answers has reached 5. (2 pts)
    • Tells them how many more correct guesses they need before the game is over. (4 pts)
• If their answer is NOT correct:
    • Let’s them know the answer is wrong. (2 pts)
```{r, eval=FALSE}
#sets 0 as the initial count
score <- 0
#gives the user 2 random numbers labeled num1 and num2 whenever the score is less
#than 5
while(score < 5){
  num1 <- sample(2:12, 1)
  num2 <- sample(2:12, 1)
  ans <- num1 * num2
  #takes in the user response using readline()
  cat(paste0("What is", num1, "x", num2, "? \n"))
  user_response <- readline()
  #if the response is correct, adds one to the correct answers count and lets
  #them know their answer is correct
  if(user_response == ans){
    score <- score + 1
    cat("Your answer is right! \n")
  #if the user's score is equal to 5, the game exits
    if(score == 5){
      break
    }
  #informs the user that they need 5 minus their current score in order to complete the game
    cat(paste0("You need ", 5 - score, "more correct guesses before the game is over. \n"))
  #if the user's response is incorrect, this lets them know
  }else{
    cat("You're answer is wrong. \n")
  }
}
```

## Question 9
Initialize a data frame named my_binom_stats with 18 rows and 2 columns. Name the first column y and the second column pdf_y. Set the first column, y, to the possible outcomes from a Binom(size=17,prob=0.83) (BE CAREFUL WITH THIS STEP!), and fill pdf_y using the appropriate stats package function.

Initialize a second, empty data frame named my_binom_loops with 18 rows and 2 columns. Name the first column y and the second column pdf_y. Then, using nested loops:
• Iterate over the 18 rows.
• Iterate over the 2 columns
• IF the inner loop is on the first column, add the appropriate number to the y column.
• OTHERWISE, fill pdf_y using the appropriate stats package function.


Finally, use the all_equal() function (from the dplyr package) to confirm that the my_binom_stats and the my_binom_loops data frames are equivalent.
```{r}
#initializes a data frame named my_binom_stats with 18 rows and 2 columns
my_binom_stats <- data.frame(matrix(nrow = 18, ncol = 2))
#names the first column y and the second column pdf_y
colnames(my_binom_stats) <- c("y", "pdf_y")
#Set the first column, y, to the possible outcomes from a Binom(size=17,prob=0.83)
my_binom_stats$y <- c(0:17)
#fills the probability density function of y
my_binom_stats$pdf_y <- dbinom(0:17, size = 17, prob = 0.83)
my_binom_stats

#initializes a data frame named my_binom_loops with 18 rows and 2 columns
my_binom_loops <- data.frame(matrix(nrow = 18, ncol = 2))
#names the first column y and the second column pdf_y
colnames(my_binom_loops) <- c("y", "pdf_y")
#iterates over the rows and columns and fulfills the given tasks
for(i in 1:18){
  for(j in 1:2){
    if(j==1){
      my_binom_loops$y = c(0:17)
    }else{
      my_binom_loops$pdf_y = dbinom(0:17, size = 17, prob = 0.83)
    }
  }
}

#uses the all_equal function to check if my_binom_stats and my_binom_loops are equivalent
all_equal(my_binom_stats, my_binom_loops)
```


## Question 10
Using set.seed(147) and the sample() function, randomly rearrange the rows of my_binom_stats from the previous question. Assign the result to my_binom_shuffle. What are the top 6 values in y?
```{r}
set.seed(147)
#randomly rearranges the rows of my_binom_stats and assigns it to my_binom_shuffle
my_binom_shuffle <- sample(nrow(my_binom_stats))
#prints my_binom_shuffle
my_binom_shuffle

#The top 6 values in "y" are 13,15,2,1,9,3
```


## Question 11
Read in the “gas.tsv” file available on Canvas. It contains the cost of filling a car with either premium or regular gas. Store the result in a variable named messy_gas. Each row represents a single car (this fact is important, you may need to modify messy_gas to account for this!!!).


Tidy the data set using pivot_longer(). Store the result in a data frame named pivot_gas.

Next, initialize an empty data frame named loop_gas that has the same column names as pivot_gas. Fill in the name and value columns with the values from messy_gas using control flow. You may fill in any other columns as you see fit.

Finally, use the all_equal() function (from the dplyr package) to confirm that the pivot_gas and the loop_gas data frames are equivalent.
```{r}
messy_gas <-
  # The file is actually CSV format!! NOTE: (3 pts)
  read_csv("gas.tsv", 
           skip = 3, # SKIP 3 LINES: NOTE: (2 pts)
           col_types = cols())

# Add a column identifying the vehicle NOTE: (2 pts)
messy_gas <- 
  messy_gas %>%
  mutate(vehicle = 1:10)
# messy_gas

# Pivot longer ONLY the columns for gas type NOTE: (3 pts)
pivot_gas <- 
  messy_gas %>%
  pivot_longer(cols = c("premium", "regular"))
# pivot_gas


# Initialize the empty data frame  
# Does NOT need to be filled with NAs NOTE: (2 pts)
loop_gas <- 
  tibble(vehicle = rep(NA_integer_, 20),
         name    = rep(NA_character_, 20),
         value   = rep(NA_real_, 20))
# loop_gas

# Initialize the index for the `loop_gas` row NOTE: (2 pts)
tidy_row <- 1

# Iterate over the rows and columns of `messy_gas`. NOTE: (2 pts)
# Since rows represent a specific car, using `car` as
# the outer loop row index makes it clear
for(car in 1:10){
  # iterate over the first two columns of `messy_gas` NOTE: (1 pt)
  for (column in 1:2){
    # Identify which car the observation belongs to NOTE: (1 pts)
    loop_gas[tidy_row, "vehicle"] <- car
    
    # conditionally label if the observation is for
    # premium or regular gar, depending on which column
    # of `messy_gas` we are reading from NOTE: (4 pts)
    if(column == 1){
      loop_gas[tidy_row, "name"] <- "premium"
    } else {
      loop_gas[tidy_row, "name"] <- "regular"
    }
    
    # Asign the price value in `messy_gas` to the `value`
    # column of `loop_gas` NOTE: (2 pts)
    loop_gas[tidy_row, "value"] <- messy_gas[car, column]
     
    # Move on to the next row of loop_gas NOTE: (1 pt)
    tidy_row <- tidy_row + 1
  }
}

# loop_gas

all_equal(loop_gas, pivot_gas)
```

