---
title: 'STAT 147 | Practice Exercises'
subtitle: 'Week 8 - Control Flow'
author: "Chirag Kulkarni (862132127)"
output:
  html_document:
    toc: true
    toc_depth: 4
    number_sections: false
    toc_float:
      collapsed: false
      smooth_scroll: false
---

```{r setup, include=FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
```


***
# Exercise 1

The definition of the **absolute value** of $x$ is:  
$$
|x| = f(x) =
\begin{cases}
-x,  & \text{if $x$ < 0} \\
\ \ x, & \text{otherwise}
\end{cases}
$$  
Write a chunk of code that uses an `if` statement to produce the absolute value.

```{r}
x <- -7
abs <- x
if (x < 0) {
    abs = -x
}
cat("The absolute value of ", x, " is ", abs , "\n" )
```

# Exercise 2
Utilize an if-else structure to return the maximum value from a numeric vector `x` of **length 2**.

```{r}
x <- c(8, 5)
if(x[1] > x[2]) {
    cat("Max value is", x[1], "\n")
} else cat("Max value is", x[2], "\n")
```

# Exercise 3
Write a for loop that iterates over the numbers 1 to 7 and prints the cube of each number using print().

```{r}
x <- 1:7
for (x in 1:7){
  print(x*x*x)
}
```

# Exercise 4
Write a while loop that prints out standard random normal numbers (use rnorm()) but stops (breaks) if you get a number bigger than 1.

```{r}
set.seed(3)
while (TRUE) {
  x <- rnorm(1)
  print(x)
  if (x > 1) {
    break
  }
}

```

# Exercise 5
Using `next`, adapt the loop from last exercise so that it doesn’t print negative numbers.

```{r}
set.seed(3)
while (TRUE) {
  x <- rnorm(1)
  if (x < 0) {
    next
  }
  print(x)
  if (x > 1) {
    break
  }
}
```

# Exercise 6
Create an empty numeric vector of length 20 named `flips`.  Using a for loop, simulate the flip of a fair coin twenty times (a single coin flip is a binomial with size = 1), keeping track of the individual outcomes (1 = heads, 0 = tails) in your pre-allocated vector, `flips`.

```{r}
x <- 20
outcome <- vector(length = x, mode = "integer")
for (i in 1:20) {
  outcome[i] <- sample(c(0L, 1L), 1)
}
outcome
```

***


<br>
<br>
<br>
<br>
<p style="font-size:30px;color:#003DA5">
<b>Congratulations!</b>  You've completed the practice exercise.  Please be sure to submit your `.Rmd` and `.html` files to Canvas.
</p>