---
title: 'STAT 147 | Practice Exercise'
subtitle: 'R Markdown'
author: "Chirag Kulkarni"
date: "`r format(Sys.time(), '2 February, 2021')`"
output:
  html_document:
    toc: yes 
    toc_float:
      collapsed: false
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```

**Acknowledgment**: Part of this R Markdown template is adapted from David Dalpiaz (UIUC) and Dr. Wenxiu Ma (UCR).


# Instructions

This week, you will first learn the basic R Markdown syntax. 

- First, download the `rmd` file `Lastname_Firstname_practice_05_.Rmd` from Canvas. It gives you an option to "open" or "save". Make sure that you **save** the file to your computer to a known location.
- Open the `rmd` file in RStudio and click **`Knit to HTML`** in the `Knit` dropdown menu to render it to HTML format. (You would need to install LaTex if you wanted to Knit to PDF).  
- Read the `rmd` file and the rendered `html` file side-by-side to see how the `html` document was generated!  
- Feel free to play around with this document! Change it. Break it. Fix it. The best way to learn R Markdown (or really almost anything) is to try, fail, then find out what you did wrong.

Next, you will review some example code. 

- Read over the code and the output. If you have any questions about certain functions or parameters, it is the time to ask!
- There are some exercises through out this document. Replace **INSERT_YOUR_ANSWER** with your own answers, knit the file to HTML, and check your results.

**Submission Guidelines**

- After you completed all exercises, save your file to `Lastname_Firstname_practice_05.Rmd` (for example, `Flores_Analisa_practice_05.Rmd`). The html file will automatically have the same name as the .Rmd document once the .Rmd file is Knitted, but with a `.html` file extension (for example, `Flores_Analisa_practice_05.html`).
- Submit your `rmd` **and** `html` files to the appropriate Canvas submission forms in this week's module. 

***

# A reminder to "FAIL FAST"

Remember learning about "failing fast"?  If not, go back and re-watch the videos because failing fast is **1 billion times more important when working with `.rmd` files!**

Any time you make a change (EVEN A MINOR CHANGE), re-compile/knit the document just to be sure that it **can** knit.  If you do too much work between knittings and you encounter a compiling issue... well, good luck with that!

***

# RMarkdown Basics

[RMarkdown](http://rmarkdown.rstudio.com/) at its core is a combination of [`R`](https://cran.r-project.org/) and [Markdown](https://daringfireball.net/projects/markdown/) used to generate reproducible reports for data analyses.

Markdown and `R` are mixed together in a `.rmd` file, which can then be rendered into a number of formats including `.html`, `.pdf`, and `.docx`. There will be a strong preference for `.html` in this course. 

Have a look at this `.rmd` file to see how this document was generated! It should be read alongside the rendered `.html` to best understand how everything works. You can also modify the `.rmd` file and re-knit along the way to see what effects your modifications have.

Formatting text is easy. **This is bold.** *This is italics.* `This text appears as monospaced.`  

# Header 1
## Header 2
### Header 3

- Unordered list element 1.
- Unordered list element 2.
- Unordered list element 3.
    - Sublist
    - Sublist
        - Sub-sublist

1. Ordered list element 1.
1. Ordered list element 2.
1. Ordered list element 3.

***

# YAML basics

"**Y**AML **A**in't **M**arkup **L**anguage"!  Instead, the YAML contains metadata that can be used to instruct R Studio on how to process the R Markdown document when you knit it.

In the rendered html document turn your attention to to the table of contents, click around a bit, scroll up and down, then come back to this [YAML basics] section.

You'll notice that the table of contents only shows the level-one headers (the headers starting with a single hashtag `#`) unless you click on it.  If there are sub-headings, they are revealed.  Also notice that if you click on a particular section header, the document animates a smooth scroll to that section.

Now, please go to the top of the .Rmd file. and look at the YAML.  **FIRST** be sure to insert your name in the space provided!!!  Next, see what happens if you change `collapsed: true` to `collapsed: false` and change `smooth_scroll: true` to `smooth_scroll: false`.

Now the table of contents should not be collapsed, and when you click on a section header in the table of contents your documents *leaps* directly to that section, with no smooth scrolling. 

Leave both `collapsed` and `smooth_scroll` as `false`.

***

# Packages

Packages are key to using `R`. The community generated packages are a large part of `R`'s success, and it is extremely rare to perform an analysis without using at least some packages. Once installed, packages must be loaded before they are used.

## Install necessary packages

```{r, echo=T}
# You only need to run the install.packages() commands once. 
# Then you can comment them out by adding `#` at the beginning of each line

# install.packages("rmarkdown", repos="http://cran.rstudio.com/")
# install.packages("yaml", repos="http://cran.rstudio.com/")
```

Note that `rmarkdown` is actually a package in `R`! If `R` never prompts you to install `rmarkdown` and its associated packages when first creating an RMarkdown document, use the above command to install them manually.

***

# `R` Code Chunks

So far we have only used Markdown to create html. This is useful by itself, but the real power of RMarkdown comes when we add `R`. There are two ways we can do this. We can use `R` code chunks, or run `R` inline.

The following is an example of an `R` code chunk.  

```{r}
# 3 random samples
sample(1:10, 5, replace = TRUE)
sample(1:10, 5, replace = TRUE)
sample(1:10, 5, replace = TRUE)
```

The function `set.seed()` fixes the way in which `R` produces random variables, so by setting the seed, any randomness will be identical at every run.

```{r}
# 3 random samples with the same seed
set.seed(147)
sample(1:10, 5, replace = TRUE)
set.seed(147)
sample(1:10, 5, replace = TRUE)
set.seed(147)
sample(1:10, 5, replace = TRUE)
```

Here is another `R` code chunk. **Please change 147 to the last 4 digits of your SID**

```{r example_chunk_name}
# generate random normals
set.seed(7863) # Please change 147 to the last 4 digits of your SID
x <- rnorm(10) # use <- for assignment instead of =
x
```

There is a lot going on in each of the above chunks.  In the `.rmd` file, notice the syntax that creates and ends the chunk. Also note that `example_chunk` is the chunk name. Everything between the start and end syntax must be valid `R` code. **Chunk names are not necessary**, but can become useful as your documents grow in size.

You can create an `R` code chunk quickly by pressing `Ctrl+Alt+i` (`Ctrl+Option+i` on a Mac).  In the below second level header (`## FIRSTNAME's own code chunks below`), change `FIRSTNAME` to your name, then create a new code chunk beneath it with the code to find and print the average of `x` from the above code chunk.  Create a second code chunk beneath it and simply print `"second code chunk, courtesy of FIRSTNAME"`.

***

## Chirag's own code chunks below

Write your code chunks between the lines below.

-------------------
```{r}
mean(x)
```

```{r}
"Second code chunk, courtesy of Chirag"
```


-------------------

When you are done, knit the document and see the result.

## The `setup` code chunk

Next, in the `.rmd` just beneath the YAML top matter, you will notice an `R` code chunk named `setup`.  This chunk IS optional, but it allows you to set the values of knitr parameters for the entire document.  

In the `setup` code chunk at the top of the `.rmd` file, change `echo = TRUE` to `echo = FALSE`, re-knit the document, and note the differences.  You shouldn't see any more actual code, but the results of that code are still being printed out.

Now change the `echo` back to `TRUE`, and then change the `eval = TRUE` to `eval = FALSE`.  Re-knit, and notice that the code is being shown, but the code is not being executed, so no results are printed.

Please **return** both `echo` and `eval` back to `TRUE`.

## Chunk Options

There are many [chunk options](http://yihui.name/knitr/options/). Above, two options were introduced: `eval` and `echo`, how they are frequently used, and how to change them **globally**.  Next, you will see how to change them **locally** for individual chunks.

```{r not_run, eval = FALSE}
?log
x
```

Using `eval = FALSE` the above chunk displays the code, but it is not run. The `?` code pulls up documentation of a function. The `x` code prints all values inside the vector x. With `eval = FALSE` (or simply `eval = F`), there is no output generated.

```{r not_displayed, echo = FALSE}
print("Hello World!")
```

Above, we see output, but no code! This is done using `echo = FALSE` (or simply, `echo=F`), which is often useful.

One final useful option is `error`.  If any code chunks have errors in them, the entire `.rmd` file will not compile/export!!!  To avoid this problem, add `error = TRUE`, allowing you to have an occasional error here and there.

```{r allow_error, echo = TRUE, error = TRUE}
# No variable `y` has been created
# so the below code should produce
# an error.
print(y)
```

```{r allow_error2, echo = TRUE, error = TRUE}
# The below code is missing a closing parenthesis
print("Hello World!")
```

## Inline `R`

`R` can also be ran in the middle of exposition. For example, the mean of the data we generated is: `r mean(x)`.

***

# Adding Math with `LaTeX`

Another benefit of RMarkdown is the ability to add [Latex](https://www.latex-project.org/about/) for mathematics typesetting. Like `R` code, there are two ways we can include Latex; displaystyle and inline.

Note that use of `LaTeX` is somewhat dependent on the resulting file format. For example, it cannot be used at all with `.docx`. To use it with `.pdf` you must have LaTeX installed on your machine.

With `.html` the `LaTeX` is not actually rendered during knitting, but actually rendered in your browser using [MathJax](https://www.mathjax.org/).

## Displaystyle `LaTeX`

Displaystyle is used for larger equations which appear centered on their own line.

$$
s = \sqrt{\frac{1}{n - 1}\sum_{i=1}^{n}(x - \bar{x})^2}
$$

## Inline LaTex

We could mix LaTeX commands in the middle of exposition, for example: $\alpha = 2$ is written in LaTeX. 

**We can actually mix `R` with Latex as well!** For example: $\bar{x} = `r mean(x)`$.

<br>
<br>

***
***

# Review

***

## Subsetting vectors

Recall that above we generated 10 random values from the normal distribution using `rnorm`, assigned the values into a vector `x`, then printed the values in vector `x`.


### **Question 1** 

Look at the output of the following code, what happens when you select with a negative index?

```{r}
x
x[2]
x[-2]
```

**ANSWER**: When you select with a negative index, it removes the index chosen and prints with the rest of x.

***

## Calculate summary statistics

Summary statistics can be calculated from the data. Here we set the option `collapse=T` for the code chunk, so that R Markdown will try to collapse all the source and output blocks from one code chunk into a single block.

```{r, collapse=T}
mean(x)
median(x)

var(x) # sample variance

n <- length(x)
sum((x-mean(x))^2)/n # population variance
mean(x^2)-(mean(x))^2 # population variance
var(x)*(n-1)/n # population variance


summary(x)
```



### **Question 2**

Change `collapse=T` to `collapse=F` and re-knit. What happens? 

**ANSWER**: Changing collapse to F doesn't allow R markdown to collapse all the source and output blocks from one code chunk into a single block.

Please **return** `collapse=F` back to `collapse=T`.



### **Question 3** 

Complete the following code to print out the maximum value and the minimum value in `x`.

**ANSWER**:

```{r}
min(x)
max(x)
```

*** 


### **Question 4** 

The `var()` function calculates the sample variance. 
$$ s^2 = \frac{1}{n-1}\sum_{i=1}^{n}(x - \bar{x})^2$$
The `sd()` function calculates the sample standard deviation. 

$$ s = \sqrt{s^2} = \sqrt{\frac{1}{n-1}\sum_{i=1}^{n}(x - \bar{x})^2}$$

Without using the `sd` function, calculate the sample standard deviation of `x`.  Your answer should be `r sd(x)`

**ANSWER**:

```{r, collapse=T}
sqrt(sum((x-mean(x))^2/(length(x)-1)))
```


*** 

### **Question 5** 

Recreate the table given in the example of [Section 16.9](https://rc2e.com/rmarkdown#recipe-static-table) of the <u>R Cookbook</u>.

**ANSWER**:

| Stooge | Year | Hair?           |
|--------|------|-----------------|
| Moe    | 1887 | Yes             |
| Larry  | 1902 | Yes             |
| Curly  | 1903 | No (ironically) |

| Stooge | Year | Hair? |
|--------|------|-----------------|
| Moe | 1887 | Yes |
| Larry | 1902 | Yes |
| Curly | 1903 | No (ironically) |

|Left   |Right | Center  | Default |
|:------|-----:|:-------:|---------|
| 12345 |12345 | 12345   | 12345   |
| text  | text | text    | text    |
| 12    | 12   | 12      | 12      |






***


<br>
<br>
<br>
<br>
<p style="font-size:30px;color:#003DA5">
<b>Congratulations!</b>  You've completed the practice exercise.  Please be sure to knit the final document and submit **BOTH** the `.Rmd` and `.html` files to Canvas.
</p>
