---
title: 'STAT 147 | Lab'
subtitle: 'Control flow | Loops, if, else, etc.'
author: "Chirag Kulkarni (862132127)"
output:
  html_document:
    toc: true
    toc_depth: 4
    number_sections: false
    toc_float:
      collapsed: false
      smooth_scroll: false
---

```{r setup, include=FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(magrittr)
```

***

# Instructions

The goal of this lab is to get practice with control flow in R. 

Please download the file `Lastname_Firstname_Lab_08.Rmd`, change the file name to reflect your own name, and complete the lab using that file.

You will also need to download the necessary data file `agility.txt`.

## A reminder to "FAIL FAST"

Any time you make a change (EVEN A MINOR CHANGE), re-compile/knit the document just to be sure that it **can** knit.

***

# Conditional statements

Decision making is an important part of programming. This can be achieved in R programming using the conditional if...else statement.

The syntax of an if statement is:

```{r, eval = F}
if (test_expression) {
  statement  # Body of `if()`
}
```

If the `test_expression` is `TRUE`, the statement gets executed. But if it’s `FALSE`, nothing happens.

Here, `test_expression` can be a logical or numeric vector, but only the first element is taken into consideration.

In the case of numeric vectors, zero is taken as `FALSE` and the rest as `TRUE.`




# Q1

1. Create a script that calculates the square root of a given integer, `x`.  If the value contained in x is negative it should return `NA`.  Do so by completing the code, replacing any instances of `...` with the appropriate code.


```{r}
x <- 25

#Determine whether the value of "x" is positive
if (x>0){
  #prints the square root
  print(sqrt(x))
  #uses the control sequence statement to define 
  #the alternative body of code in case the 1st
  #condition is not satisfied
}else{
  print("NA")
  }
```



# Q2

2. Create an R script that returns the max value of a vector x with length 3. Do so by completing the below code, replacing any instances of `...` with the appropriate code.

To do this, we need to confirm one of the following:

> If x[1] > x[2] AND x[1] > x[3], then x[1] is the maximum  
> If x[2] > x[1] AND x[2] > x[3], then x[2] is the maximum  
> If x[3] > x[1] AND x[3] > x[2], then x[3] is the maximum

Also, note that if the first two scenarios are not satisfied then the last scenario has to be true, so it does NOT need to be checked explicitly!


```{r}
#creates an R script that returns the max value of a vector x with length 3
x <- c(22,33,11)

cat("The maximum is:\n")
if(x[1] > x[2] & x[1] > x[3]){
  print(x[1])
}else if (x[2] > x[1] & x[2] > x[3]){
  print(x[2])
}else if (x[3] > x[1] & x[3] > x[2]){
  print(x[3])
}
```


***

# While Loop

A while loop is used when you want to perform a task indefinitely, until a particular condition is met. It’s a condition-controlled loop. 


The loop will continue until the condition is FALSE. 


```{r}
x = 5

# If statement is true, keep running the loop 
while (x != 0 ) {
  print(x)
  x = x - 1
}
```


If the condition is false at the start, **the while loop will never be executed at all**.

```{r}
x = 0

# If statement starts as TRUE,  the loop will never run 
while (x != 0 ) {
  print(x)
  x = x - 1
}
```



# Q3

3. Use a while loop to simulate one stock price path starting at 100. Assume the price jump occurs randomly according to a normal distribution with mean 0 and standard deviation of 0.1 each period. How long does it take to reach above 150 or below 50?

This question is a bit of a head scratcher, but let's try anyway.  First, the primary question is **How long does it take to reach above 150 or below 50?** so we should create a counter to keep track of the number of iterations.

```{r}
cntr <- 0
```

Next, we need to set the initial price to \$100.

```{r}
price <- 100
```

We need to create a while loop that will terminate if the price exceeds \$150 **OR** drops below \$50. Be sure that the initial value of `price` satisfies the conditions so that we can enter the while loop at all.

```{r, eval = F}
while(50 < price & price < 150){
  n <- n + 1
  price <- price*(1 + rnorm(mean = 0, sd = 0.01, n = 1))
}
n
```

Lastly, we need to write code in the body of the loop to add a zero mean random value to the price, and count the number of times we do this. Set the seed to `147` for repeatability.

```{r}
cntr <- 0
price <- 100

set.seed(147)
while(50 < price & price < 150){
  price <- price + rnorm(1, mean = 0, sd = 0.1)
  cntr <- cntr + 1
}

cntr
price
```

***

# Break

The `break` statement is used to exit the loop immediately. It simply jumps out of the loop altogether, and the program continues after the loop.

```{r}
x = 5

#If the statement starts as TRUE, the loop will never run
while(x != 0){
  print(x)
  x = x-1

  if(x==2){
    print("Entered IF statement, stop loop")
    break
  }
}
```

If not given an adequate stopping criteria or break statement **the loop will continue forever**. For example, if the above example started at `x = -2`. 


# Q4

4. Repeat question 3, but change the standard deviation to `0.01`, and add code to `break` if the number of iterations exceeds 1 million.

```{r}
cntr <- 0
price <- 100

set.seed(147)
  while(50 < price & price < 150){
   price <- price + rnorm(1, mean = 0, sd = 0.01)
   cntr <- cntr + 1
  if(cntr > 1000000){
    break
  }
}
cntr
price
```

Does the process exceed 1 million iterations?  Yes or No?

**ANSWER:** Yes

Report the final number of iterations and the price.

**ANSWER:** Iterations: 68412070
            Price: 150.0017
            
# Q5

5. Please google how to stop an infinite loop in R Studio.  Then, run the example code from the beginning of the [Break] section (the first example) with `x` starting at `-2`.  Run the code chunk from the Rmd file.  Please *respond with how you stopped the infinite loop*.<br><br>
NOTE:  Leave the code chunk option `eval` equal to `FALSE`.

```{r, eval = F}
x = -2

#If the statement starts as TRUE, the loop will never run
while(x != 0){
  print(x)
  x = x-1

  if(x==2){
    print("Entered IF statement, stop loop")
    break
  }
}
```

**Explanation:** 
I stopped the infinite loop by clicking the "stop" sign in the bottom left console.

***


# Repeat Loops

This loop is similar to the while loop, but it is made so that the block of instructions is executed **at least once**, no matter what the result of the condition.  `repeat` loops require an `if()` statement defining the condition, and contains a `break` to end the loop.


```{r}
response <- 42

repeat{
  response <- as.integer(response);
  if (response == 42){
    print("Well Done!");
    break
  }else{
    print("Sorry, the answer to whatever the question MUST be 42")
    print(paste0("Current answer: ", response))
    if (response < 42){
      response <- response + 1
    }else{
      response <- response - 1
    }
    print(paste0("Next answer: ", response))
    }
}
```

# Q6

6. What is the MINIMUM number of times the above loop will execute (for any possible specified value of x)?

**ANSWER:** 1

***

# For Loops

The `for` statement in `R` is a bit different from what you usually use in other programming languages. Rather than iterating over a numeric progression, `R`’s `for` statement iterates over the items of a vector or a list. The items are iterated in the order that they appear in the vector.

```{r}
colors = c("red", "green", "blue", "yellow")
#Iterate through a vector
for(x in colors){
  print(x)
}
```

```{r}
lst = list(3.14, "Hi", c(1,2,3))
#Iterate through a list
for (i in lst) {
  print(i)
}
```

If you need to execute a group of statements for a specified number of times, use sequence operator : or built-in function `seq()`.

```{r}
#Print "hello" three times
for (x in 1:3){
  print("Hello")
}
```


```{r}
#Print the square of 2,4,6,8
for (x in seq(from = 2, to = 8, by = 2)){
  print(x^2)
}
```


Like the while loop, the `break` statement is used to exit the loop immediately. It simply jumps out of the loop altogether, and the program continues after the loop.

```{r}
colors = c("red", "green", "blue", "yellow")
#Iterate through a vector
for(x in colors){
  if (x == "blue"){
    break
  }
  print(x)
}
```

NOTE: For loops do not have the same risk of "running forever", like while loops have. 


# Q7

7. Create an R script that returns the number of values that are larger than the mean of a vector.  Report the correct number. Set the seed to `147` for repeatability.

```{r}
#set the random seed for reproducibility
set.seed(147)
#generate 50 random normal random variables with mean 15 and std dev 3
x <- rnorm(50, mean = 15, sd = 3)
#calculate the sample mean of "x"
sample_avg <- mean(x)

#initialize the counter variable
cntr <- 0

#iterate through the observation in "x"
for(obs in x){
  #if the current observation is larger than the sample mean
  #count it by adding 1 to 'cntr'
  if(obs > sample_avg) cntr <- cntr + 1
}

#print the final count
cntr
```

***

# Initializing variables

Suppose you were asked to create a vector that contains the numbers 1 through 10 using loops.  Some people might take the following approach.

```{r}
my_nums <- 1
for(i in 2:10){
    my_nums <- c(my_nums, i)
}
my_nums
```

However, this is bad practice.  Technically, this solution is correct!  But the way in which the computer manages memory for a process like this is FAR from optimal.

Computers store objects in memory, and every time the dimensions of an object are extended/changed, R needs to make a temporary copy.


Instead, let's see what happens when we pre-allocate a vector 

```{r}
my_nums <- numeric(10)

for(i in 1:10){
    my_nums[i] <- i
}
my_nums
```
There is NO copying, so the calculations can be done MUCH faster!


# Q8

8. Generate the first 12 Fibonacci numbers using a **for** loop.  Note that the Fibonacci sequence is formed by starting with the number 0, 1 and then adding the last two numbers to get the next number: 0, 1, 1, 2, 3, 5, 8, etc.

```{r}
#generate the first "n" fibonacci numbers

#determine how many
n <- 12
#initialize the fib sequence 
fib = rep(0, n)
fib[2] <- 1

for(index in 3:length(fib)){
  #calculate the next fibonacci number
  next_number = fib[c(index-2)] + fib[index-1]
  
  #add the new number to the vector at the current location/index
  fib[index] = next_number
}
print(fib)
```

What is the last number in the printed sequence?

**ANSWER:** 89

***

# Reading user input

The function `readline` can be used to prompt a user to enter a value in the console, which can then be assigned to a variable of your choosing! 

For instance, the following example asks the user to "Guess a number".  When the user enters a number, the result is assigned to the variable named `user_input`.

```{r take_input, eval = F, class.source = "nocopy"}
user_input <- readline("Guess a number: ")
user_input
```



# Q9

9. Implement a simple version of a "Guess the number" game using a `repeat` loop. The user should guess a number between 1 and 10 (you can use `readline("Guess a number between 1 and 10: \n")` to get user input). If the user guesses 5, the loop should print "Good guess!!!!" and then break.  Otherwise, the loop should print "Try again!!!" and re-enter the loop.
<br><br>
**NOTE:** It's important that for anything involving `readline()`, you set the R code chunk option `eval` equal to `FALSE`.  If you don't, R will *never finish knitting your document* because it is waiting for someone to enter a value in the console.

```{r, eval = F}
#Initialize the repeat loop
repeat{
  user_guess <- readline("Guess a number between 1 and 10: ")
  if (user_guess == 5){
    print("Good guess!!!")
    break
  }else{
    print("Try again")
  }
}
```

***

# Nested loops

Let's find all of the prime numbers less than 60.  First, let's count how many there are.

To review, a [prime number](https://en.wikipedia.org/wiki/Prime_number) is a natural number greater than 1 that is not a product of two smaller natural numbers.  Therefore, 0 and 1 cannot be prime, and 2 MUST be prime.  All other numbers greater than 2 need to satisfy the aforementioned definition, that is, they are not divisible by any number between 2 and itself.

We will accomplish this doing the following:

- Loop through the number 3 through 60 as candidate prime numbers.  
- For each candidate prime number, we will loop through numbers, 2 through the number 1 less than the current number, by which we will divide the candidate prime number.  
- We will then check to see what the remainder of each division.  If at any time the remainder is zero, the number is clearly not prime and we can halt dividing the current candidate prime number.
- If, however, we reach the end of the list of divisors, then we can say the number is prime, so add one to our prime counter.

```{r}
#Since we start with one prime number (2) we start with the counter at 1
n_prime <- 1

#The remaining candidate prime numbers are 3 through 60
for(candidate_prime in 3:60){
  #For each candidate prime number, we will loop 2 through the number 
  #1 less than the current number, by which we will divide the 
  #candidate prime number
  for(divisor in 2:(candidate_prime-1)){
    #checks the remainder of each division
    remainder <- candidate_prime %% divisor
    #If at any time the remainder is zero, the number 
    #is clearly not prime and we can halt dividing the current candidate
    #prime number
    if(remainder == 0){
      #halt dividing
      break
      
      #if we reach the end of the list of divisors, then we can
      #say the number was prime, so add one to our prime counter
    }else if(divisor == (candidate_prime -1)){
      #add one to our prime counter
      n_prime <- n_prime + 1
    }
  }
}
#print to check
n_prime
```

# Q10

10. How many prime number are less than 60?

**ANSWER:** 17


***

Suppose you are asked to build a vector containing the first 20 prime numbers.  What kind of loop is best for this problem?  **While** you may be tempted to go with only **for** loops (since you know the number of prime numbers is 20). However, we **don't** know how far up the list of natural numbers we'll need to go, so we'll need a **while** loop for that! 

We will accomplish this doing the following:

- We will initialize an empty vector, `my_primes`, with 20 slots.  This is used to store the prime numbers when we find them. 

- We will store the value of 2 as the first prime number in our `my_primes` vector.

- We will then initialize TWO counters: <br>  • The first will be the `candidate_prime`, which we will add `1` to *indefinitely* until we finally reach 20 prime numbers. <br>  • The next will be `n_prime`, the variable that record how many primes we've identified, and which serves as the index used to store any prime numbers we discover.

- While the number of identified prime numbers is less than 20:  
• For each candidate prime number, we will loop the number `2` through `candidate_prime - 1` (1 less than the current prime candidate). We will divide the candidate prime number by each of these.  
• We will then check the remainder of each division.  If at any time the remainder is zero, the number is clearly not prime and we can halt dividing the current candidate prime number.  
• However, if we reach the end of the list of divisors, then we can say the number was prime, so add one to our prime counter, print out the candidate prime, and store the prime number in the vector `my_primes` at the appropriate index, `n_prime`.


```{r}
#initialize the vector in which to store the first 20 prime numbers
my_primes <- rep(NA, 20)
#store the value of 2 as the first prime number in our "my_primes" vector
my_primes[1] <- 2
#print to check
my_primes
#initialize 'candidate prime', which we will add to '1' *indefinitely*
#until we finally reach 20 prime numbers
candidate_prime <- 3
#initialize the variable that records how many primes we've identified,
#and which serves as the index used to store any prime numbers we've discovered
#into the vector 'my_primes'
n_prime <- 1
while(n_prime < 20){
  for(divisor in 2:(candidate_prime-1)){
    remainder <- candidate_prime %% divisor
    if (remainder == 0){
      break
      #if we reach the end of the list of divisors, then we can
      #say the number was prime, so add one to our prime counter,
      #then store the prime number at that location
    }else if (divisor == (candidate_prime - 1)){
      n_prime <- n_prime + 1
      my_primes[n_prime] <- candidate_prime
    }
  }
  #move on to the next candidate prime number
  candidate_prime <- candidate_prime + 1 
}
#print to check
print(my_primes)
```

# Q11

11. What is the last prime number in the list?

**ANSWER:** 71


***

# Tidying data using for loops

Let's use nested `for` loops to *tidy* up some data.  Please read in the file `agility.txt` (you may need to change the path). Convert the result to a matrix. Assign the result to a variable named `dirty_dogs`.


```{r, eval = T}
# Change "eval" to "TRUE"
dirty_dogs <- 
  # Read in the file, AND THEN...
  read_table2("agility.txt", skip = 1) %>%
  # Convert the result to a matrix
  as.matrix()

# Print to check
print(dirty_dogs)
```


Next, we are going to initialize a dataframe that will be identical to the following:

```{r, eval=T}
# Change "eval" to "TRUE"
pivot_dogs<-
  dirty_dogs %>% 
  as.data.frame() %>%
  pivot_longer( cols = c("Cody", "Dusty", "Shadow"),
             names_to = "Dog_name",
             values_to = "Time")
pivot_dogs
```

```{r}
tidy_dogs1 <-
  #initialize to empty vectors of length 
  tibble(Dog_name = character(24),
         Time = numeric(24))
```


Next we are going to use nested for-loops to fill in the `tidy_dogs1` data frame using the info inside of `dirty_dogs`.  We will do this twice, first iterating <u>rows then columns</u>, followed by <u>columns then row</u>.

**ROWS then columns:**

```{r, class.source = "nocopy"}
#initialize the row of the receiving data frame to be the first row
tidy_row <- 1
#Loop through all of the rows in the matrix
for(dirty_row in 1:8){
  #For each row of the matrix, loop through the columns
  for(dirty_col in 1:3){
    #use a control flow sequence to label the dog name
    #based on the current column
    if (dirty_col == 1){
      tidy_dogs1$Dog_name[tidy_row] <- "Cody"
    }else if (dirty_col==2){
      tidy_dogs1$Dog_name[tidy_row] <- "Dusty"
    }else{
      tidy_dogs1$Dog_name[tidy_row] <- "Shadow"
    }
    #Next, store number from the matrix row/column combo in the receiving data frame
    tidy_dogs1$Time[tidy_row] <- dirty_dogs[dirty_row, dirty_col]
    #Move on to the next row in the receiving data frame
    tidy_row <- tidy_row + 1
  #end the inner loop
  }
#end the outer loop
}
tidy_dogs1
```

# Q12

12. Is the data frame `tidy_dogs1` equivalent to `pivot_dogs`?  You *may* need to sort both data frames by name and time to confirm.

```{r}
#checks if tidy_dogs1 are equivalent to pivot_dogs
all.equal(tidy_dogs1, pivot_dogs)
```


**NOW YOU TRY!**  **COLUMNS then rows**

Redo the previous example. Initialize a new data frame or tibble named `tidy_dogs2` and reset `tidy_row` to `1`.  Leave the rest of the code untouched, except swap the contents of the for-loop iterators (that is, place `dirty_col in 1:3` in the outer for loop, and `dirty_row in 1:8` in the nested for loop.)  Run the code. 

```{r}
#initialize the row of the receiving data frame to be the first row
tidy_row <- 1
#Loop through all of the rows in the matrix
for(dirty_col in 1:3){
  #For each row of the matrix, loop through the columns
  for(dirty_row in 1:8){
    #use a control flow sequence to label the dog name
    #based on the current column
    if (dirty_col == 1){
      tidy_dogs1$Dog_name[tidy_row] <- "Cody"
    }else if (dirty_col==2){
      tidy_dogs1$Dog_name[tidy_row] <- "Dusty"
    }else{
      tidy_dogs1$Dog_name[tidy_row] <- "Shadow"
    }
    #Next, store number from the matrix row/column combo in the receiving data frame
    tidy_dogs1$Time[tidy_row] <- dirty_dogs[dirty_row, dirty_col]
    #Move on to the next row in the receiving data frame
    tidy_row <- tidy_row + 1
  #end the inner loop
  }
#end the outer loop
}
tidy_dogs1
```

# Q13

13. Are the two data frames `tidy_dogs1` and `tidy_dogs2` equivalent?  You may need to sort both data frames by name and time to confirm.

```{r}
# WRITE CODE TO CHECK
```

# Q14

14.  Which dog has the fastest course time?  

```{r}
max(tidy_dogs1$Time)
```

**ANSWER:** Shadow

****


<br>
<br>
<br>
<br>
<p style="font-size:30px;color:#003DA5">
<b>Congratulations!</b>  You've completed the lab.  Please be sure to submit your answers to Canvas. You will also need to submit your `.Rmd` and `.html` files to Canvas.
</p>
