---
title: 'STAT 147 | Lab'
subtitle: 'Week 7 - R Probabilities'
author: "Chirag Kulkarni (862132127)"
output:
  html_document:
    highlight: pygments
    theme: cerulean
    toc: true
    toc_depth: 4
    number_sections: false
    toc_float:
      collapsed: true
      smooth_scroll: false
mainfont: FiraSans
---

```{r setup, include=FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(magrittr)
library(knitr)
```


***
# Instructions

The goal of this lab is to become comfortable with random variables, their probabilities, and their simulation.

Please download the file `Lastname_Firstname_Lab_07.Rmd`, change the file name to reflect your own name, and complete the lab using that file.

## A reminder to "FAIL FAST"

Any time you make a change (EVEN A MINOR CHANGE), re-compile/knit the document just to be sure that it **can** knit.

***

# Probability Distributions

**RECALL:**

  • For random variables, **probability density** = **probability mass function** = $P(X = x)$.  
  
  • For random variables, **survival function** =  $P(X > q) = 1 - P(X \leq q)$  and $P(X \geq a) = 1 - P(X < a)$  
  
  • For **discrete** random variables only, $P(X < q + 1) = P(X \leq q)$.  
  
  • For **continuous** random variables only, $P(X \leq q) = P(X < q)$ and $P(X \geq a) = P(X > a)$.  

<br>

**NOTE**:  One can  use the PDF/PMF, CDF and SDF (CDF with tail.prob = FALSE) functions in R to calculate probability density (mass) functions, cumulative distribution functions  [$P(X \leq x)$], and survival distribution functions [$P(X > x)$]. The quantile function can be used to generate percentiles.

***

## Question 1

1. Suppose X is a continuous random variable that represents the life length (in months) of a battery. Also suppose X is **normally distributed** with a **mean** of **48** months and a **standard deviation** of **6** months.

### Q1, Part (i)

 \ \ \ \ (i) Find the probability that a battery selected at random will last between 36 and 60
(**inclusively**)  months. 

We want to use R to compute $P(36 \leq X \leq  60)$.

NOTE: $P(36 \leq X \leq 60) = P(X \leq 60) - P(X < 36)$. 

We write the following R code:

```{r}
# P(36 <= X <= 60)
# Central Region = Big Left Tail - Small Left Tail
#     = P(X <= 60) - P(X < 36)
pnorm(60, mean = 48, sd = 6) - pnorm(36, mean = 48, sd = 6)
```

### Q1, Part (ii)

 \ \ \ \  (ii) Find the probability that a battery selected at random will last more than 63 months. 

```{r}
# P(X > 63)
## method 1
pnorm(63, mean = 48, sd = 6,
      lower.tail = FALSE)
## method 2
# P(X > 63) = 1 - P(X <= 63)
1-pnorm(63, mean = 48, sd = 6)
```


### Q1, Part (iii)

The lecture videos cover dxxx(), pxxx(), and rxxx().  However, the quantile/percentile function `qxxx()` was not discussed.  Please use the help documentation and google as needed to explore the function and it's meaning. 

 \ \ \ \ (iii) Find the 90th percentile for the life length of the battery. i.e., Find x such that P (X ≤ x) = 0.90.

```{r}
#finds the 90th percentile for the life length of the battery
qnorm(p = 0.90, mean = 60, sd = 6)
```

Thus, for a normal density function with a mean of 60 and a standard deviation of 6, the value of $x$ that gives 90% of the area under the curve to its left is `r qnorm(p = 0.90, mean = 60, sd = 6)`. In other words, 90% of these batteries will have a life length of `r qnorm(p = 0.90, mean = 60, sd = 6)` months or less.

***


## Question 2

2. Suppose the probability that a person who regularly watches college football is female with a probability of **0.45**. **Twenty** college football watchers are selected at random.

Unlike question 1, the distribution of the random variable is not directly specified. However, the distribution can be determined based on the setting.  

**20** people are selected at random, i.e. **20 trials**. The outcome of each trial can be categorized as a "success" or a "failure". Each trial has a **"success" probability of 0.45**.  The random response is the number of successes out of the 20 independent trials.  Therefore, one can ascertain that the random variable follows a **$Binomial(n=20, p = 0.45)$** distribution.

Now, please consider the wording in the following questions to determine the correct way to answer them:

### Q2, Part (i)

\ \ \ \ (i) What is the probability that **exactly** 9 are female?  

```{r}
# Exact probabilities are PMFs (Probability Mass Functions)
# P(X = 9)
dbinom(9, size = 20, prob = 0.45)

# method 2
# P(X = 9) = P (X <= 9) - P(X <= 8)
pbinom(9, size = 20, prob = 0.45) - pbinom(8, size = 20, prob = 0.45)
```

### Q2, Part (ii)

\ \ \ \ (ii) What is the probability that at **most** 11 are female?  

```{r}
# P(X <= 11)
pbinom(11, size = 20, prob = 0.45)
```

### Q2, Part (iii)

\ \ \ \ (iii) What is the probability that more than 10 are female? 

```{r}
# Method 1
# P(X > 10)
pbinom(10, size = 20, prob = 0.45, lower.tail = FALSE)

# Method 2
# P(X > 10) = 1 - P(X <= 10)
1 - pbinom(10, size = 20, prob = 0.45)
```

### Q2, Part (iv)

**NOW YOU TRY!!!**

\ \ \ \ (iv) What is the probability that at least 9 are female?  

```{r}
# P(X >= 9) = 1 - P(X < 9)
# = 1 - P(X <= 8)
1 - pbinom(8, size = 20, prob = 0.45)
```


### Q2, Part (v)

\ \ \ \ (v) What is the probability that between, and including, 8 and 12 are female?

```{r}
# P(8 <= X <= 12)
# = P(X <= 12) - P(X < 8)
# = P(X <= 12) - P(X <= 7)
pbinom(12, size = 20, prob = 0.45) - pbinom(7, size = 20, prob = 0.45)
```

Let's get creative!  Let's calculate the individual exact probabilities, then add them up. Replace the `...` as necessary.

**When you have made the correction, please change the `eval = F` to `eval = T` in the code chunk options.**

```{r, eval=T}
outcome_set <- 8:12
# Print to check
outcome_set

# Calculate the exact probabilities
exact_probs <- dbinom(outcome_set, size = 20, prob = 0.45)
# Print to check
names(exact_probs) <- c("P(X=8)", "P(X=9)", "P(X=10)", "P(X=11)", "P(X=12)")
# PRINT TO CHECK
exact_probs

# Add them up
central_interval <- sum(exact_probs)
names(central_interval) <- "P(8 <= X <= 12)"
# PRINT TO CHECK
central_interval
```

***


## Question 3

3. Suppose X is a discrete random variable such that X ∼ Bin(x; n = 10, p = 0.50). Use R to generate the pdf and cdf of X.

```{r}
#creates a pdf and cdf of X such that X ∼ Bin(x; n = 10, p = 0.50)
binom_distn <-
  tibble(x = 0:10)
binom_distn
```

### Q3, Part (i)

\ \ \ \ (i) Let's use the mutate function to add a new column for the exact probabilities for each observation.  Please modify the following, replacing the `...` with the correct code.  

**When you have made the correction, please change the `eval = F` to `eval = T` in the code chunk options.**

```{r, eval = T}
#adds a new column for the exact probabilities for each observation 
binom_distn <-
  binom_distn %>%
    mutate(PMF = dbinom(x, size = 10, prob = 0.5))
binom_distn
```


As a sanity check, let's confirm that the probabilities sum to 1.  Use the `summarise()` function to sum the `PMF` column.

```{r}
#confirms that the probabilities sum to 1
binom_distn %>%
  summarise(total_prob = sum(PMF))
```

### Q3, Part (ii)

\ \ \ \ (ii) Next, let's use the mutate function to add a new column for the cumulative probabilities for each observation.  Please modify the following, replacing the `...` with the correct code.  

**When you have made the correction, please change the `eval = F` to `eval = T` in the code chunk options.**

```{r, eval = F}
#adds a new column for the cumulative probabilities for each observation 
binom_distn <-
  binom_distn %>%
   mutate(CDF = pbinom(x, size = 10, prob = 0.5))
binom_distn
```


### Q3, Part (iii)

\ \ \ \ (iii) Create a plot of x vs the pmf of x.

```{r}
#uses the plot function to create a plot of x vs PMF
binom_distn %$%
  plot(x, PMF,
       main = "x vs P(X = x)",
       col = "blue",
       type = "o",
       pch = 18,
       cex = 1.5
       )
```

### Q3, Part (iv)

\ \ \ \ (iv) **Now you try!**   Create a plot of the cumulative distribution, i.e., `x` vs `P(X <= x)`.  Change the color of the plot to purple.

```{r}
binom_distn %$%
  plot(x, binom_distn$CDF,
       main = "x vs P(X <= x)",
       col = "purple",
       type = "o",
       pch = 18,
       cex = 1.5
       )
```

***

## Question 4

4. Dylan believes that 30% of tablet pc owners prefer the *Microsoft Surface Pro*. To test his claim, he selects a random sample of 25 tablet pc owners. Let X = # of owners that selected the *Microsoft Surface Pro* as their preference. Use this information to complete the following using R.

### Q4, Part (i)

\ \ \ \ (i) What is the probability that **exactly 9** owners selected the *Microsoft Surface Pro*?

```{r}
# P(X = 9)
# Use dbinom (x, size = sample size, prob = probability of success)
exactly_9 <- dbinom(x = 9, size = 25, prob = 0.3)
#print the answer 
exactly_9
```

### Q4, Part (ii)

\ \ \ \ (ii) What is the probability that **no more than 9** owners selected the *Microsoft Surface Pro*?

```{r}
# P(X <= 9)
# Use pbinom (q, size = sample size, prob = probability of success)
no_more_than_9 <- pbinom(q = 9, size = 25, prob = 0.3)
#prints the answer
no_more_than_9
```

### Q4, Part (iii)

\ \ \ \ (iii) What is the probability that **9 or more** owners selected the *Microsoft Surface Pro*? 

```{r}
# P(X >= 9)
# = P(X > 8)
# 1 - P(X <= 8)
# Use pbinom (q, size = sample size, prob = probability of success)
nine_or_more <- pbinom(q=8, size=25, prob = 0.30,
                       lower.tail = FALSE)
#prints the answer
nine_or_more
```

### Q4, Part (iv)

\ \ \ \ (iv) What is the probability that between 9 and 12 owners, inclusively, selected the *Microsoft Surface Pro* as their preference?

```{r}
# P(9 <= X <= 12) = P(X <= 12) - P(X < 9) = P(X <= 12) - P(X <= 8)
ge9_and_lel2 <- pbinom(12, size = 25, prob = 0.30) - pbinom(8, size = 25, prob = 0.30)
#prints the answer
ge9_and_lel2
```

### Q4, Part (v)

\ \ \ \ (v)  What is the probability that less than 7 or more than 12 owners selected the *Microsoft Surface Pro*? Be sure to write your R commands and the results of each step in the space provided.

```{r}
# P(X < 7 OR X > 12)
# method 1: Add two tails
# P(X < 7 OR X > 12)
# = P(X < 7) + P(X > 12)
# = P(X <= 6) + P(X > 12)
lt7_or_gt12 <-
  pbinom(6, size = 25, prob = 0.30) + pbinom(12, size = 25, prob = 0.30, lower.tail = FALSE)
#print the answer
lt7_or_gt12

# method 2: Remove central region
# P(X < 7 OR X > 12)
# = 1 - P(7 <= X <= 12)
# = 1 - central region
# = 1 - {P(X <= 12) - P(X <= 6)}
central_region <- pbinom(12, size = 25, prob = 0.30) - pbinom(6, size = 25, prob = 0.30)
1 - central_region
```

****

## Question 5

5. Microsoft claims that the Surface Pro 4 has a battery life that is **normally distributed** with a **mean** of 8 hours (when web browsing) and a **standard deviation** of 1.25 hours. Mark wants to buy a Surface Pro 4 for work. He goes to a local store and a salesperson selects a Surface Pro 4 at random for him. Mark takes the Surface Pro 4 to his office, fully charges the battery and then uses it until it runs out of power. Let X = the battery life of the Surface Pro 4.

**NOTE:** The above question assumes $X \sim N(\mu,\sigma)$.  
In `R`,  

 use `dnorm(x,mean,sd)` to generate $P(X = x)$ = pdf of $X$  
 use `pnorm(q,mean,sd)` to generate $P(X \leq x)$ = cdf of $X$  
 use `pnorm(q,mean,sd,lower.tail = FALSE)` to generate $P(X > x)$ = 1 - cdf of $X$   


### Q5, Part (i)

\ \ \ \ (i) Find the probability that the battery life  is  less than 6.4 hours?

```{r}
# P(X < 6.4)
# use pnorm(x,mean,sd)
lower1 <- pnorm(6.4, mean = 8, sd = 1.25)
# print the answer
lower1
```

### Q5, Part (ii)

\ \ \ \ (ii) Find the probability that the battery life  is  greater than 6.4 hours?

```{r}
# P(X > 6.4)
# use pnorm(x, mu, sigma, lower = FALSE) to get upper tail
upper1 <- pnorm(6.4, 8, 1.25, lower.tail = FALSE)
#print the answer
upper1
```

### Q5, Part (iii)

\ \ \ \ (iii) Confirm that lower1 and upper1 are complements of one another (i.e. confirm that they sum to 1).

```{r}
#confirms that lower1 and upper1 are complements of one another
sum(lower1, upper1)
```


### Q5, Part (iv)

\ \ \ \ (iv) Find the probability that the battery life  is  between 6.5 and 8.5 hours.

```{r}
#finds the probability that the battery life is between 6.5 and 8.5 hours
battery <-  pnorm(8.5, mean = 8, sd = 1.25)- pnorm(6.5,8,1.25)
battery
```



***

## Question 6

6. Refer to question 5.  

### Q6, Part (i)

\ \ \ \ (i) Find the 75th percentile (quantile) for battery life. i.e., find x such that $P(X \leq x) = 0.75$.

```{r}
#use qnorm(quantile, mean, sd) to generate 95th percentile
quantile75 <- qnorm(p = 0.75, mean = 8, sd = 1.25)
#print the answer
quantile75
```

### Q6, Part (ii)

\ \ \ \ (ii) Find the 87th percentile (quantile). i.e., find x such that $P(X \leq x) = 0.87$.

```{r}
#use qnorm(quantile, mean, sd) to generate 87th percentile
quantile87 <- qnorm(p = 0.87, mean = 8, sd = 1.25)
#print the answer
quantile87
```

### Q6, Part (iii)

\ \ \ \ (iii) **ON YOUR OWN!!!**  Using what you know about complements and quantiles/percentiles, find $x$ such that $P(X > x) = 0.40$.

```{r}
#p(X > x) = 1 - p(X <= x)
#use qnorm(quantile, mean, sd) to generate 40th percentile
quantile40 <- 1- qnorm(p = 0.40, mean = 8, sd = 1.25)
#print the answer
quantile40

```

***


<br>
<br>
<br>
<br>
<p style="font-size:30px;color:#003DA5">
<b>Congratulations!</b>  You've completed the lab.  Please be sure to submit your answers to Canvas. You will also need to submit your `.Rmd` and `.html` files to Canvas.
</p>
