---
title: "Lab 10"
author: "Chirag Kulkarni"
date: "3/9/2021"
output: 
  html_document:
    toc: true
    toc_depth: 4
    number_sections: false
    toc_float:
      collapsed: false
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1
Write your own bisection method function which takes in as input five arguments: FUN which is the function to be analyzed, lo which is the lower bound of the zero crossing, hi which is the upper bound of the zero cross, tol which is the tolerated distance between lo and hi for the loop to terminate, and max_itr which is the maximum number of iterations allowed before termination with an error.
```{r}
#pick a name for the function
my_bisect <-

  #use the 'function()' keyword to assign the variable a function definition
  function(FUN, lo = -100, hi = 100, tol = 1e-2, max_itr = 1e3){
    f <- FUN

    #lo <- 0
    #hi <- 15
    # Define the iterator
    itr <- 0
    # calculate the difference
    delta <- abs(hi - lo)
    # Begin the while loop
    # Define the tolerance (0.01) and the max.itr (10^5)
    repeat{
        #Calculate the midpoint of `a` and `b`
        mid <- (lo + hi)/2
        if(f(mid) > 0){
            # If f(mid) > 0, then we know the solution 
            # for `x`` such that f(x) = 0 lies between
            # `a` and `mid`, so change `b` to `mid`
            hi <- mid
        } else {
            # Otherwise, change `a` to `mid`
            lo <- mid
        }
        # Calculate the difference of the upper
        # and lower bound to check in the while
        # loop condition
        delta <- abs(hi - lo)
        itr <- itr + 1
        if(itr > 10^5){
            cat("Error:  Maximum number of iterations reached")
            break
        } else  if (delta > max_itr) {
            break
        }
    }
    return(mid)
  }
my_fun <- function(x) 1/(1+exp(-(x-17/3))) - 1/2
#call the function to check that it works
#you can pass an 'anonymous' function to the 'FUN' argument, as below
#or assign the numeric function to a variable, and pass it to the 'FUN' argument after
my_bisect(FUN = my_fun, lo = 0, hi = 15, tol = 1)
my_bisect(FUN = my_fun, lo = 0, hi = 15, tol = 1, max_itr = 2)
```


## Question 2
Refer to the first question. Try creating a new, monotonic (strictly increasing or decreasing) function and pass it to the FUN argument of the my_bisect() function! A straight line satisfies this criteria.
```{r}
#pick a name for the function
my_bisect <-

  #use the 'function()' keyword to assign the variable a function definition
  function(FUN, lo = -100, hi = 100, tol = 1e-2, max_itr = 1e3){
    f <- FUN

    #lo <- 0
    #hi <- 15
    # Define the iterator
    itr <- 0
    # calculate the difference
    delta <- abs(hi - lo)
    # Begin the while loop
    # Define the tolerance (0.01) and the max.itr (10^5)
    repeat{
        #Calculate the midpoint of `a` and `b`
        mid <- (lo + hi)/2
        if(f(mid) > 0){
            # If f(mid) > 0, then we know the solution 
            # for `x`` such that f(x) = 0 lies between
            # `a` and `mid`, so change `b` to `mid`
            hi <- mid
        } else {
            # Otherwise, change `a` to `mid`
            lo <- mid
        }
        # Calculate the difference of the upper
        # and lower bound to check in the while
        # loop condition
        delta <- abs(hi - lo)
        itr <- itr + 1
        if(itr > 10^5){
            cat("Error:  Maximum number of iterations reached")
            break
        } else  if (delta > max_itr) {
            break
        }
    }
    return(mid)
  }
my_fun <- function(x) 1/(1+exp(-(x-17/3))) - 1/2
#call the function to check that it works
#you can pass an 'anonymous' function to the 'FUN' argument, as below
#or assign the numeric function to a variable, and pass it to the 'FUN' argument after
my_bisect(FUN = my_fun, lo = 0, hi = 15, tol = 1)
my_bisect(FUN = my_fun, lo = 0, hi = 15, tol = 1, max_itr = 2)
```


## Question 3
Refer to the Monty Carlo approximation of π from this week’s practice exercise. We will do the same thing, except this time we will increment the sample size n until we finally get an approximation of π such that, when we round to 2 digits, the estimate equals 3.14.

We will also use the plus_one() and calc_dist() function from this week’s practice exercise.
```{r}
              #PART 1
calc_dist <- function (radius = 0.5) {
  # Assign variables
  lo = -radius
  hi = radius
  # Generate a bivariate point (x,y)
  point <- runif(2, min = lo, max = hi)
  # Calculate distance
  distance <- sqrt(sum(point^2))
  # RETURN the final value
  return(distance)
}
plus_one <- function(itr){
  # Add one to the input
  return(itr + 1)
}

            #PART 2
total_points <- 500
circle_pts <- 0
for(i in 1:total_points){
  distance <- calc_dist(radius = 0.5)
  if(distance <= 0.5) circle_pts <- plus_one(circle_pts)
}
my_pi <- 4 * circle_pts / total_points
my_pi

           #PART 3
set.seed(147)

#initialize
total_points <- 500
circle_pts <- 0

#use a repeat loop so that we enter the loop AT LEAST ONCE
repeat{
  #for each iteration through the loop:
  #record the total number of simulated points
  total_points <- plus_one(circle_pts)
  #calculate the distance of the new point from the origin
  distance <- calc_dist(radius = 0.5)
  #if the distance is small enough, count it as being within the circle
  if(distance <= 0.05) circle_pts <- plus_one(circle_pts)
  #calculate the estimate of pi
  my_pi <- 4 * circle_pts/total_points
  #round the above estimate to 2 digits
  round_pi <- round(my_pi, digits = 2)
  #stop iterating if the rounded estimate satisfies our condition
  if(round_pi == 3.14) break
  #add a "safety" break to protect against infinite loops
  if(total_points >= 10000) break
}
total_points
```


## Question 4
Incorporate the above loop into a function named estimate_pi(). Do not include the set.seed(147). BE CAREFUL! You may be tempted to use total_points and circle_pts as arguments in the “formals” of the function, but this is not necessary. Simply write a function with no arguments which approximates π. Have it return a NAMED VECTOR, with the first entry equal to my_pi and the second equal to the number of iteration to get that pi approximation. For instance

  c(my_pi 3.140956,  total_points209)


Then, when you have it working successfully, call the following code and report the output.
```{r}
#vector <-
  
```


## Question 5
Confidence intervals: Follow along with the following exercise and answer the subquestions along the way.
```{r}
load("ames.RData")

#set the gross liveable area as the population
population <- ames$Gr.Liv.Area
#draw a sample of 60 houses
samp <- sample(population, 60)


hist(samp)

sample_mean <- mean(samp)

se <- sd(samp)/sqrt(60)
lower <- sample_mean - 1.96 * se
upper <- sample_mean + 1.96 * se
conf_intvl_95 <- c(lower, upper)
conf_intvl_95

mu <- mean(population)
mu

#define how many times you want to simulate drawing a sample 
k <- 50
#initialize storaze vectors for the associated mean and sd of each simulation
samp_mean <- rep(NA, k)
samp_sd <- rep(NA, k)
#define the size of each individual sample
n <- 60

for(i in 1:k){
  samp <- sample(population, n) # obtain a sample of size n = 60 from the population
  samp_mean[i] <- mean(samp)    # save sample mean in ith element of samp_mean
  samp_sd[i] <- sd(samp)        # save sample sd in ith element of samp_sd
}

lower_vector <- samp_mean - 1.96 * samp_sd / sqrt(n) 
upper_vector <- samp_mean + 1.96 * samp_sd / sqrt(n)

c(lower_vector[1], upper_vector[1])
```


## Question 6 
Confidence interval visualization with a function
```{r}
plot_ci(lo = lower_vector, hi = upper_vector, m = mean(population))
```

## Question 7
```{r}
plot_ci <- 
    function(lo, hi, m){
      k <- 50
    # Adjust the graph parameters    
    par(# margin size (mar)
        mar=c(2, 1, 1, 1),
        # axis label locations (mgp)
        mgp=c(2.7, 0.7, 0))
    
    # Bind the vectors into columns of a matrix
    lo_hi <- cbind(-1*lo, hi)
    
    # Calculate the ranges
    ranges <- rowSums(lo_hi)
    
    # Identify the largest range
    ci.max <- max(ranges)
    
    # Specify the lowest and highest possible x values for the plot
    xR <- m + ci.max*c(-1, 1)
    # Specify the lowest and highest possible y values for the plot
    yR <- c(0, 41*k/40)
    
    # Create a blank plot with the above specified dimentions 
    plot(xR, yR, type='n', xlab='', ylab='', axes=FALSE)
    # Add a vertical dashed line above the true mean
    abline(v=m, lty=2, col='gray')
    # Add a label for the true mean on the x-axis
    axis(1, at=m, paste("mu = ",round(m,4)), cex.axis=1.15)
    #axis(2)
    # plot all k intervals
    for(i in 1:k){
        # Calculate the midpoint of the interval
        midpoint_i <- x <- mean(c(hi[i],lo[i]))
        # Isolate the conf interval for the ith sample
        ci <- c(lo[i],hi[i])
        
        # ONLY IF the confidence interval DOES NOT contain the 
        # true mean within it, add a red line
        if(!contains(lo[i],hi[i],m)){
            
            # Add a red circle around the midpoint of the C.I.
            # #F05133 is the color hex code for the shade of red I like
            points(midpoint_i, i, cex=1.4, col="red")
            
            # Add a think red line accross the interval
            # #F05133 is the color hex code for the shade of red I like
            lines(ci, rep(i, 2), col="#F05133", lwd=5)
        }
        
        # Always add a black interval for every C.I.
        # Start with a solid black dot
        points(x, i, pch=20, cex=1.2, col=1)
        # End with a black line covering the region
        lines(ci, rep(i, 2), col=1)
    }
}
```


## Question 8
Type I error (α)
```{r}
#PART 1
mu <- mean(population)
sigma <- sd(population)
std_err <- sigma/sqrt(60)

x <- seq(mu - 3 * sigma, mu + 3 * sigma, 1)

plot(x, dnorm(x, mean = mu, sd = std_err), type = "l", col = "blue")
lines(x, dnorm(x, mean = mu, sd = sigma), type = "l")

#PART 2
alpha <- 0.05

lower_crit_val <- qnorm(alpha/2, mean = mu, sd = std_err)
upper_crit_val <- qnorm(alpha/2, mean = mu, sd = std_err, lower.tail = FALSE)
lower_crit_val
upper_crit_val

#PART 3
k <- 50
test_record <- rep(NA, k)
n <- 60

for (i in 1:k){
  samp <- sample(population, n)
  test_result <- t.test(samp, mu = 1499.69)
  test_record[i] <- test_result$p.value < 0.05
}
sum(test_record) / length(test_record)
```

